%% TO DO: Monotonicity

%% TO DO: more rules for convenience, e.g. defining the \bot symbol,
%% and then defining in and out rules for it

%% Prove that the $\vee$elimination rule is necessary
%% if we didn't have it, then our system would be sound relative to valuations

%% Show that $\phi\vee\bot \equiv \phi$

%% TO DO: find my old exercises involving "adoring somebody" etc

%% TO DO: synonyms for "entails" or "implies"

%% TO DO: translation from predicate symbols to English

%% TO DO: discuss that -> may not be truth-functional

%% TO DO: serial relation = entire relation

\documentclass[fleqn]{tufte-book}
\setcounter{secnumdepth}{0}
% \includeonly{rules}
\title{How Logic Works}
\author{Hans Halvorson}
\input{preamble.tex}
\renewcommand{\emph}{\textbf}
\usepackage{xcolor,colortbl}
\definecolor{Gray}{gray}{0.85}
\usepackage[acronym]{glossaries-extra}
\usepackage{svg}
\makeglossaries
\usepackage{venndiagram}
\usepackage{enumitem}
\usepackage{mathtools}
\usepackage{amsthm}
\usepackage{amssymb}
\usepackage{tcolorbox}
\usepackage{multirow}
\newtheorem*{prop*}{Proposition}
\newtheorem{prop}{Proposition}
\numberwithin{prop}{chapter}
\newtheorem*{lemma*}{Lemma}
\newtheorem*{fct}{Finite completeness theorem}
\newtheorem*{dnf}{DNF theorem}
\newtheorem*{subthm}{Substitution theorem}
\newtheorem*{sothm}{Soundness theorem}
\theoremstyle{definition}
\newtheorem*{defn}{Definition}
\newtheorem*{example}{Example}
\newtheorem{exercise}{Exercise}
\newtheorem{exercises}[exercise]{Exercise}
\numberwithin{exercise}{chapter}
% \publisher{Princeton University Press}
\usepackage{array}
\usepackage[caption=false]{subfig}
\usepackage{tikz}
\usetikzlibrary{calc,trees,positioning,arrows,fit,shapes,calc}
\newcommand{\mymk}[1]{%
  \tikz[baseline=(char.base)]\node[anchor=south west, draw,rectangle, rounded corners, inner sep=2pt, minimum size=7mm,
    text height=2mm](char){#1} ;}
\newcommand*\circled[1]{\tikz[baseline=(char.base)]{
            \node[shape=circle,draw,inner sep=2pt] (char) {#1};}}
\usepackage[framemethod=TikZ]{mdframed}
\newcounter{axi}\setcounter{axi}{0}
\renewcommand{\theaxi}{\arabic{axi}}
\newenvironment{law}[2][]{%
    \refstepcounter{axi}
 \ifstrempty{#1}%
% if condition (without title)
{\mdfsetup{%
    frametitle={%
        \tikz[baseline=(current bounding box.east),outer sep=0pt]
        \node[anchor=east,rectangle,fill=blue!20]
        {\strut Axiom~\theaxi};}
    }%
% else condition (with title)
}{\mdfsetup{%
    frametitle={%
        \tikz[baseline=(current bounding box.east),outer sep=0pt]
        \node[anchor=east,rectangle,fill=blue!20]
        {\strut ~#1};}%
    }%
}%
% Both conditions
\mdfsetup{%
    skipabove=2em,
    skipbelow=2em,
    innertopmargin=10pt,linecolor=blue!20,%
    linewidth=2pt,topline=true,%
    frametitleaboveskip=\dimexpr-\ht\strutbox\relax%
}
\begin{mdframed}[]\relax}{%
\end{mdframed}}
\usepackage{tikz-cd}
% \setcounter{secnumdepth}{1}
% \setcounter{tocdepth}{0}
\newcommand{\lra}{\leftrightarrow}

\newtcbox{irule}[1]{coltitle=blue!20!black, colback=black!1!white, 
  colframe=black!10!white,title=#1}

% \tbcset{coltitle=blue!20!black, colback=black!1!white, 
% colframe=black!10!white}

% \tcbset{fonttitle=\bfseries,nobeforeafter,center title}

% \usepackage{imakeidx}

% \makeindex

\tcbuselibrary{skins,xparse}

% \includeonly{models}

\makeglossaries


\doublespacing
\begin{document}
\frontmatter

\maketitle

\tableofcontents

\newglossaryentry{valid}
{
  name={valid},
  description={An argument is valid just in case its premises provide
    decisive support for its conclusion, alternatively, if the truth
    of its premises guarantees the truth of its conclusion} }

\newglossaryentry{atomic sentence} { name={atomic sentence},
  first={\emph{atomic sentence}},
  description={An atomic sentence is a sentence that does not have any
    other sentence as a proper syntactic part.  In propositional
    logic, atomic sentences are represented by capital letters such as
    $P,Q,R,\dots $.  In quantifier logic, the atomic sentences are
    either relation symbols applied to closed terms, such as $Rab$, or equalities between closed terms, such as $a=b$.}
}


\newglossaryentry{interpretation}
{
  name={interpretation},
  description={An interpretation is an assignment of symbols to set-theoretic structures.}
}


\chapter{Preface}

This book has no subject matter --- or, to be more precise, it's about
both everything and nothing.  For the science of logic has no
doctrines or creeds.  There is no set of beliefs that distinguishes
the logical people from the non-logical people, not the beliefs of the
European Enlightenment, nor the deliverances of contemporary natural
science, nor the opinions of some Princeton philosopher.  Simply put,
learning the science of logic can't be reduced to learning any
particular facts at all; it's learning a skill, namely, the skill to
discern between good and bad arguments.

There are numerous reasons why you need this skill, no matter what you
end up doing with your life.  First, being logical will help you
reason about how to get what you want.  Second, many of our society's
best jobs require strong logic skills --- whether it be programming
computers, buying stock options, curing diseases, prosecuting
criminals, discovering alternative energy sources, or interpreting the
constitution.  (And that's not even to speak of that all-important
task of parenting and raising intellectually healthy children.)
Third, regardless of your ideological bent --- whether you're
religious, atheist, or agnostic --- you surely want to do everything
you can to ensure that your view aligns with reality.  For this task,
logic is an invaluable tool, if only to protect you from the plethora
of bad arguments you'll hear in your life.

Our goal here is no more or less than to initiate you into the most
up-to-date account of what makes an argument good.  We'll give you the
tools, but it's up to you to decide how you're going to use them.

\medskip {\it {\bf Note} about how to use this book:} The first six
chapters of this book (up through ``quantifying'') correspond roughly
to a first course in formal logic at an average university in the
United States.  We cover the material in about a quarter of the number
of pages of some well-known logic textbooks, which was intentional:
logic should organize your already existing thoughts, not add to the
jumble of thoughts that need to be organized.

In Chapter 7, we gently ramp up to applications of formal logic, and
in particular to the application of formal logic to the study of
itself (i.e.\ logical metatheory).  In one sense, this material goes
beyond a typical first course in logic, but we would like to change
that.  This material is not intrinsically more difficult than what has
gone before.  In fact, what can make it seem difficult is that it's
less clear what rules of reasoning one is permitted to use.  However,
we try to explain clearly that the rules of reasoning here are none
other than those that were learned in the previous chapters.

Having shown how formal logic can be used to formulate theories, we
then begin to use formal logic to formulate a theory about itself.  In
Chapter 8, we use the theory of sets to define the notion of an
interpretation of a language, and Tarski's rigorous definition of {\it
  truth}.  In Chapter 9, we present a theory about propositional
logic, proving both soundness and completeness theorems.  Finally, in
Chapter 10, we sketch the outlines of a similar theory about
quantifier logic.  Thus, the chapters form a natural staircase leading
up to the treasure-chest of advanced applications of logic, including
modal logic, set theory, and G{\"o}del's theorem.

\medskip {\it {\bf Note} to the instructor, or to the person deciding
  which logic book to use:} To express a thought, or to make an
argument, or to formulate a theory, you have to pick a language to
use.  Similarly, to become more logical, you have to adopt some
particular system of logic.  I've made a choice for you here, and here
is my rationale.

The first choice point is between ``trees'' and ``arguments.''  The
advantage of trees is that they are really easy, and require little
mental exertion.  But wait, isn't the goal to become the most
excellent logical thinkers that we can be?  People don't become better
logical thinkers by being taught a recipe that somebody else found,
and that lets them effectively turn their minds off.  But that's
precisely the point of logical trees: they give students a {\it
  recipe} for evaluating simple arguments.  If the goal were to
manufacture an army of logical automata, then I might well use trees.
But since I'm teaching human beings, I prefer to teach them that
distinctively human skill of making and evaluating arguments.
Accordingly, this book focuses primarily on how to make rigorous
arguments, and secondarily on the (non-algorithmic) skill of detecting
bad arguments.

The second choice point is whether to represent arguments using a
``Fitch-style'' or a ``Lemmon-style'' system.  This is a difficult
decision.  Fitch-style is very intuitive and has a shallow learning
curve.  Unfortunately, Fitch-style argument is opaque to reflection:
it's difficult for students to see why Fitch-style works, and even
more difficult for them to imagine how the rules might have been
different (thereby reinforcing an unfortunate perception that there is
no human creativity involved in formulating the laws of logic).  In
contrast, Lemmon-style argument has a steeper learning curve, but it's
more flexible and transparent to reflection.  It's relatively easy to
see why the system works, and it's easy to tweak the rules and see how
things would come out different.  So, the steeper learning curve of
the Lemmon approach is a price I'm willing to pay to enable my reader
to become a more clear and creative logical thinker.

There are at least three ways in which this book differs from many
other introductory logic books.  First, we teach intrinsic methods
(e.g.\ reasoning) prior to teaching extrinsic methods (e.g.\ checking
with truth tables).  Here our philosophical convictions have been
reinforced by our pedagogical experience, viz.\ that an
argument-centric approach leads to students gaining a higher level of
mastery and confidence, and that the skills gained are more easily
transferable to other intellectual tasks.

Second, and for similar reasons, we treat formal validity as defined
in terms of the inference rules, and not in terms of
truth-preservation.  Our issue here is that a precise notion of
truth-preservation requires a previous understanding of valid
reasoning in set theory.  So, we want to be forthright with our reader
that we, at least, don't have a clear and distinct perception of what
truth-preservation amounts to, at least not until we explicate the
notion in terms of set theory.  By taking formal validity as defined
by the rules, we emphasize that genuine creativity is required to
design rules that both capture and sharpen our intuitions about
validity.

Third, in Chapter \ref{theories}, we ease students away from rigid
formal proofs, and towards ``informal rigor.''  Again, the goal is for
the students to be able to transfer their knowledge to other
disciplines where rigid formal proofs might not be appropriate or
useful.  Fourth, in Chapter \ref{meta}, we introduce logical
metatheory as ``just another theory'' by explaining how to formulate
the metatheory for propositional with predicate logic and a bit of set
theory.  Thus, we close the circle of logic back upon itself.

\medskip {\it {\bf Note} to the reader:} We put an asterisk to the
left of exercises that might be perceived as more difficult than the
preceding ones.  In the case of proofs, it often occurs that a
difficult result is basically a version of the law of excluded middle:
$P\vee \neg P$.  So, if you get stuck, you might try first to prove
excluded middle, and then to use that to get the result you want.

\mainmatter

\newglossaryentry{double turnstile}% label
{
  name={\ensuremath{\vDash}},% default display
  description={semantic implication: whenever the sentences to the
    left of the double turnstile are true, the sentence to the right
    of the double turnstile is true},% description
  category=symbol% category label
}

\newglossaryentry{turnstile}% label
{
  name={\ensuremath{\vdash}},% default display
  description={the relation of provability defined by the inference rules},% description
  category=symbol% category label
} 

\newglossaryentry{main column}
{
  name={main column},
  description={is the column in the truth table of a sentence corresponding to the main connective of that sentence}
}

\newglossaryentry{soundness theorem}
{
  name={main column},
  description={shows that provable sequents are truth-preserving}
} 

\newglossaryentry{substitution instance}
{
  name={substitution instance}, first={\emph{substitution instance}},
  description={A substitution instance of a sentence is any other
    sentence that results from the first by a uniform replacement of
    non-logical terms.  i.e.\ it is any sentence that could result
    from translating that sentence to another language.}
}

\newglossaryentry{main connective} { name={main connective},
  first={\emph{main connective}}, description={for a propositional
    logic sentence $\phi$, the last connective in the construction of
    $\phi$}}

\newglossaryentry{substitution} { name={substitution},
  first={\emph{substitution}}, description={when some non-logical
    symbols are replaced with other suitable syntactic structures}}

\newglossaryentry{subformula} { name={subformula},
  first={\emph{subformula}}, description={a subformula of $\phi$ is
    any formula that occurs in the construction of $\phi$}}

\newglossaryentry{reconstrual} { name={reconstrual}, description={A
    reconstrual assigns non-logical symbols to corresponding syntactic
    structures.  For example, a reconstrual assigns a predicate symbol
    to a formula with one free variable.}  }

\newglossaryentry{dependency number} { name={dependency number},
  first={\emph{dependency number}}, description={a number in the
    leftmost column of a proof that displays which assumptions are in
    force at a particular step in the proof}  }

\newglossaryentry{translation}
{
  name={translation},
  description={A translation is a map from formulas to formulas,
    generated by a reconstrual of the non-logical vocabulary in those formulas.}
}

\newglossaryentry{model}
{
  name={model},
  description={A model of a theory is an interpretation in which all the theory's sentences are true.}
}

\newglossaryentry{sequent}
{
  name={sequent},
  description={A sequent consists of a list of sentences (premises), a
    turnstile $\vdash$, and another sentence (conclusion).  It is the
    symbolic representation of a valid argument form}
}

\newglossaryentry{disjunction}
{
  name={disjunction},
  description={A disjunction is a sentence whose main connective is
    ``or'', symbolized by $\vee$}
}

\newglossaryentry{conditional}
{
  name = {conditional},
  description={A conditional sentence is one whose main connective is
    ``if\dots then'', symbolized by $\to$}
}

\newglossaryentry{extension}
{
  name={extension},
  description={The extension of a relation is the set of things that
    stand in that relation.  The extension of a name is the individual that it denotes}
}

\newglossaryentry{intension}
{
  name={intension},
  description={The intension of a relation is, intuitively, what the relation means}
}

\newglossaryentry{formula}
{
  name={formula},
  description={A formula (alternately, well-formed formula) is a
    string of symbols that could become a sentence by the addition of quantifiers}
}

\newglossaryentry{antecedent}
{
  name={antecedent},
  description={The antecedent of a conditional is the sentence that
    occurs after ``if'', e.g.\ in ``if it is raining then the sidewalk
    is wet,'' the sentence ``it is raining'' is the antecedent}}

\newglossaryentry{consequent}
{
  name={consequent},
  description={The consequent of a conditional is the sentence that
    occurs after ``then'', e.g.\ in ``if it is raining then the sidewalk is
    wet'' the sentence ``the sidewalk is wet'' is the consequent}}

\newglossaryentry{literal}
{
  name={literal},
  description={A propositional logic sentence is called a literal if
    it is atomic or negated atomic}}

\newglossaryentry{truth-functional}
{
  name={truth-functional},
  description={A connective is truth-functional just in case its truth
    value is a function of the truth value of the relevant component
    sentences}}

\newglossaryentry{signature}
{
  name={signature},
  description={A signature is a set of non-logical symbols:
    propositional constants, relation, function, and constant symbols}}

\newglossaryentry{consistent} { name={consistent}, description={A
    collection of sentences is semantically consistent if they can all
    be true simultaneously.  A collection of sentences is
    proof-theoretically consistent if no contradiction can be derived
    from them} }

\newglossaryentry{expressively complete} { name={expressively
    complete}, description={A collection of connectives is
    expressively complete if it can express all truth functions} }

\newglossaryentry{biconditional} { name={biconditional},
  description={A biconditional is a statement of the form ``$\phi$ if
    and only if $\psi$.'' It asserts that $\phi$ is a necessary and
    sufficient condition for $\psi$} }

\newglossaryentry{sufficient condition} { name={sufficient condition},
  description={In a conditional statement ``if $\phi$ then $\psi$'',
    the antecedent $\phi$ is a sufficient condition for $\psi$} }

\newglossaryentry{necessary condition} { name={necessary condition},
  description={In a conditional statement ``if $\phi$ then $\psi$'',
    the consequent $\psi$ is a necessary condition for $\phi$} }

\newglossaryentry{inconsistency} { name={inconsistency},
  description={An inconsistency is a sentence that is false in all
    situations} }

\newglossaryentry{contingency} { name={contingency},
  description={A contingency is a sentence that is true in some
    situations and false in other situations}}

\newglossaryentry{tautology} { name={tautology}, description={A
    tautology is a sentence $\phi$ that is provable from no premises,
    i.e.\ $\vdash \phi$, or that is true merely in virtue of its form,
    or that is true in all situations}}

\newglossaryentry{counterexample}
{
  name={counterexample},
  description={A counterexample to an argument is a formalization of the notion of a situation, or state of affairs,
    in which the premises are true and the conclusion is false.} }

\newglossaryentry{universal quantifier}{ name={universal quantifier},
  first={\emph{universal quantifier}}, description={the symbol
    $\forall$ that plays the role of ``all'' or ``every''}}

\newglossaryentry{existential quantifier}{ name={existential quantifier},
  first={\emph{existential quantifier}}, description={the symbol
    $\exists$ that plays the role of ``some'' or ``there is''}}

\newglossaryentry{variable}{ name={variable}, first={\emph{variable}}, description={a symbol such as $x$, which
    plays the role of an open term}}



\newglossaryentry{dnf}{type=\acronymtype, name={DNF},
  description={disjunctive normal form}, first={\textbf{disjunctive normal
      form (DNF)}}}

\newglossaryentry{cnf}{type=\acronymtype, name={CNF},
  description={conjunctive normal form}, first={\textbf{conjunctive
      normal form (CNF)}}}

\newglossaryentry{em}{type=\acronymtype, name={EM},
  description={law of excluded middle}, first={\textbf{law of excluded
      middle (EM)}}}

\newglossaryentry{raa}{type=\acronymtype, name={RAA},
  description={reductio ad absurdum}, first={\emph{reductio ad
      absurdum (RAA)}}}

\newglossaryentry{dm}{type=\acronymtype, name={DM},
  description={De Morgan's rule}, first={\textbf{DeMorgan's rule (DM)}}}

\newglossaryentry{efq}{type=\acronymtype, name={EFQ},
  description={ex falso quodlibet}, first={\emph{ex falso quodlibet
      (EFQ)}}}

\newglossaryentry{qn}{type=\acronymtype, name={QN},
  description={quantifier negation equivalences},
  first={\emph{quantifier negation equivalences (QN)}}}

\newglossaryentry{complete}{name={complete},
  first={\textbf{completeness}}, description={(1) A complete proof system
    is one the proves everything it should: if $\phi\vDash\psi$ then
    $\phi\vdash\psi$. (2) A theory $T$ is complete just in case
    $T\vdash\phi$ or $T\vdash\neg\phi$, for every sentence $\phi$ in
    its language.}}

\newglossaryentry{sound}{name={sound}, first={\textbf{soundness}},
  description={A sound proof system is one that doesn't prove things
    it shouldn't: if $\phi\vdash\psi$ then $\phi\vDash\psi$.}}

 



\include{lfh}

\include{deducing}

\include{supposing}

\include{building}

\include{truth}

\include{quantifying}

\include{theorizing}

\include{models}

\include{meta}

\include{meta-pred}

\include{limit}

\backmatter

\appendix

\singlespacing

\include{rules}

\include{derived}

\include{tables}

\include{algorithms}





% \include{equiv}



%% basic set theory


%% de-interpretation

%% Glossary
%% Conjunction
%% Main Connective
%% Truth-value
%% Well formed formula
%% Fallacy


% \newglossaryentry{idempotent}
% {
%   name={idempotent},
%   description={If $\circ$ is a binary function, then an idempotent is
%     an $x$ such that $x\circ x=x$} }

% \newglossaryentry{quantifier}
% {
%   name={quantifier},
%   description={A phrase to express how many of a certain kind of thing
%     exists, in our symbolic language, either $\exists$ or $\forall$} }

% \newglossaryentry{scope}
% {
%   name={scope},
%   description={The scope of a quantifier in a formula is the
%     subformula to which the quantifier is applied} }


% \newglossaryentry{truth value}
% {
%   name={truth value},
%   description={is a mathematical object (in particular, either $0$ or $1$) that can be assigned to a
%     sentence, representing its status as either true or false}
% }

% \newglossaryentry{fallacy}
% {
%   name={fallacy},
%   description={Technically speaking, any invalid argument is a
%     fallacy.  Historically speaking, fallacies are invalid
%     arguments that people have often mistaken for valid}
% }




% \newglossaryentry{first-order logic}
% { name={first-order logic},
%   description={A first-order logic is one in which quantifiers apply
%     only to elements, and not to predicates.  In other words, a
%     first-order logic cannot say, ``There is a property $P$ such that
%     \dots ''}}


% \newglossaryentry{disjunctive normal form}
% {
%   name={disjunctive normal form},
%   description={A formula is in disjunctive normal form just in case it
%     is a disjunction of conjunctions of literals}
% }

% \newglossaryentry{main connective}
% {
%   name={main connective},
%   description={the main connective of a formula is the final
%     connective applied in the construction of that formula}
% }

% \newglossaryentry{interpretation}
% {
%   name={interpretation},
%   description={An interpretation of a signature $\Sigma$ is an
%     assignment of a domain set $M$, and of extensions for all
%     relation and function symbols.  The extensions of relation symbols
%     are subsets of Cartesian products $M\times\cdots\times M$.}
% }

% \newglossaryentry{empty set}
% {
%   name={empty set \ensuremath{\emptyset}},
%   description={The empty set $\emptyset$ is the unique set that contains no elements}
% }




% \newglossaryentry{bot}
% {
%   name={\ensuremath{\bot}},
%   description={the false: a sentence that is always false, and which
%     implies every other sentence},
%   sort=bot
% }

% \newglossaryentry{top}
% {
%   name={\ensuremath{\top}},
%   description={the true: a sentence that is always true, and which
%     is implied by any other sentence}, 
%   sort=top
% }

% \newglossaryentry{contradiction} { name={contradiction}, description={A
%     contradiction, in the strict sense, is a formula of the form
%     $\phi\wedge\neg\phi$.  A contradiction, in the loose sense, is a
%     formula that is always false} }

% \newglossaryentry{term} { name={term}, description={Semantically
%     speaking, a (closed) term is a symbol that denotes an object.
%     Syntactically speaking, the terms of a predicate logic signature
%     are defined inductively: names and variables are terms, and a
%     function symbols applied to terms yields another term}}


% \newglossaryentry{subformula} { name={subformula}, description={A
%     subformula of a formula is any formula that occurs in the
%     construction tree of that formula}}

% \newglossaryentry{sentence} { name={sentence}, description={In
%     predicate logic, a sentence is a formula with no free variables}}

% \newglossaryentry{free variable} { name={free variable},
%   description={A variable is free in a formula if it is not bound by
%     a quantifier}}

% \newglossaryentry{valuation} { name={valuation},
%   description={A valuation of a language is a special case of a translation, 
%     where the atomic sentences of the language are assigned truth-values.  A
%     valuation corresponds to a row in a truth-table}}

% \newglossaryentry{truth-table} { name={truth-table},
%   description={A truth-table is a systematic listing of all
%     truth-values assigned to a formula and its subformulas}}



%% subformula

%% substitution instance

% \glsaddall
\printglossaries


% \printindex


\end{document}



%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% End:
